Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-10-22 analysis/paper/paper.qmd
editing methods

In addition to the metric attributes from previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used additional morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded were: (1) the tip of the point; (2) the left side of mid-point; (3) the left wing, which can be overlapped with the mid-point; (4) the left side of the mid-point on the tang curve; (5) the left side of 5mm above the basal end; (6) the center of the basal end; (7) the tip of the basal end, which can be overlapped with the center of the basal end ; (8) the right side of 5mm above the basal end ; (9) the right side of the mid-point on the tang curve; (10) the right wing; and (11) the right side of mid-pint, which can be overlapped with the right wing. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks and export them as XY coordinate data (@tbl-landmark-abbreviations).

@@

To measure the variations between stemmed points, we used Coefficients of Variation (CV) with confidence intervals. CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes including distinguishing different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2% (i.e. CV for planned production = 1.7 %) and about 60% (i.e. CV for random production = 57.7%) and the values over the upper limit likely reflect intended attempts to differentiate the products. Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea, we sat roughly 25 as the threshold to decide the type of transmission biases.

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Since our sample size is 152 artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results. [@kelley2007sample; @albatineh2014confidence].

We chose Principal Component Analysis (PCA) to compute Pearson's r correlations between the attributes as previous studies did [@bettinger1999point; @eerkens2008cultural; @garvey2018current] and to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points. PCA is often used to detect relevant information from dataset, indicate relationships between variables as well as cases, extract general patterns in data, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life]. PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].

All data treatment, analyses and visualization were performed in the R environment [@core2021team]. Our R code and data including the raw images are openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].

>>>>

In addition to the metric attributes from previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we also used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artefacts, and export them as XY coordinate data for further analysis.

@@

To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products. Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea, we selected roughly 25 as the threshold to decide between guided variation (\<25) and indirect bias (\>25). I THINK WE NEED TO EXPLAIN THIS DECISION OF THE THRESHOLD VALUE IN MORE DETAIL. I EXPECT REVIEWERS WILL HAVE QUESTIONS ABOUT WHY 25 IS THE NUMBER YOU CHOSE.

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Since our sample size is `r total_number_of_sp` artifacts, we used the standard CV formula rather than that of @vanpool2011quantitative, and we computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results. [@kelley2007sample; @albatineh2014confidence]. I THINK THIS 'standard CV formula' IS A BAD CHOICE AND CONFUSING TO REVIEWERS. THIS IS BECAUSE IN SOME OF OUR PLOTS BELOW THERE ARE AS FEW AS 2 ARTEFACTS, THAT IS A VERY SMALL SAMPLE SIZE. GIVEN WHAT YOU HAVE WRITTEN ABOVE, WE SHOULD USE THE 'CORRECTED CV' BECAUSE MANY OF OUR COMPARISONS ARE WITH N\<25, FOR EXAMPLE, MOST SITES HAVE \<25 SP, SO WE *DO* HAVE SMALL SAMPLES AND WE NEED TO DEAL WITH IT USING THE BEST METHODS AVAILABLE.

To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we chose Principal Component Analysis (PCA). PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points. PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life]. PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].

All data treatment, analyses and visualization were performed in the R environment [@core2021team]. Our R code and data, including the original artefact images, are fully and openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].
